---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Unzip data file and read it as data frame 'activity'

```{r, results = 'hide'}
unzip("activity.zip")
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date)
```

Take a look at 'activity'
```{r}
head(activity)
```

## What is mean total number of steps taken per day?

Aggregate data for number of steps taken each day

```{r, results = 'hide', message = FALSE, warning = FALSE}
library(dplyr)
selected_data <- activity %>%
        select(date, steps)
agg_data <- selected_data %>%
        group_by(date) %>%
                summarise(total_steps = sum(steps))
```

### Barplot of number of steps taken each day

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
ggplot(data = agg_data, aes(x = date, y = total_steps)) +
        geom_bar(stat = "identity", fill = "cyan", color = "black", alpha = 0.5) +
                labs(x = "Date", y = "Steps", title = "Number of steps taken by day") + 
                        theme(plot.title = element_text(hjust = 0.5))
```

### Average number of steps for each day

```{r, results = 'hide'}
selected_data <- activity %>%
        select(date, steps)
agg_data <- selected_data %>%
        group_by(date) %>%
                summarise(mean_steps = mean(steps, na.rm = TRUE), median_steps = median(steps, na.rm = TRUE))
agg_data$date <- as.character(agg_data$date)
```

Print as a table

```{r, message = FALSE, results = 'asis'}
library(xtable)
print(xtable(agg_data), type = "html")
```

## What is the average daily activity pattern?

Aggregate average number of steps per day

```{r, results = 'hide'}
selected_data <- activity %>%
        select(date, steps)
agg_data <- selected_data %>%
        group_by(date) %>%
                summarise(mean_steps = mean(steps, na.rm = TRUE))
```

Plot time series
             
```{r, message = FALSE, warning = FALSE}
ggplot(data = agg_data, aes(x = date, y = mean_steps)) +
        geom_line(color = "blue") +
                labs(x = "Date", y = "Average number of steps", title = "Time series of average number of steps") +
                        theme(plot.title = element_text(hjust = 0.5))
``` 

### The 5-minute interval that, on average, contains the maximum number of steps

```{r}
selected_data <- activity %>%
        filter(steps == max(steps, na.rm = TRUE)) %>%
                select(date, interval)
selected_data
```

## Imputing missing values

```{r}
imputed_data <- activity[complete.cases(activity), ]
```

### Histogram of the total number of steps taken each day after missing values are imputed

Aggregate data for number of steps taken each day

```{r}
imputed_data$date <- as.character(imputed_data$date)
selected_data <- imputed_data %>%
        select(date, steps)
agg_data <- selected_data %>%
        group_by(date) %>%
                summarise(total_steps = sum(steps))
```

Barplot of number of steps taken each day

```{r, message = FALSE, warning = FALSE}
ggplot(data = agg_data, aes(x = date, y = total_steps)) +
        geom_bar(stat = "identity", fill = "cyan", color = "black", alpha = 0.5) +
                labs(x = "Date", y = "Steps", title = "Number of steps taken by day w/o NAs") + 
                        theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_blank())
```

## Are there differences in activity patterns between weekdays and weekends?

```{r, message = FALSE, warning = FALSE}
wdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
wends <- c("Saturday", "Sunday")

wday_activity <- activity %>%
        filter(weekdays(date) %in% wdays)

wend_activity <- activity %>%
        filter(weekdays(date) %in% wends)

selected_wday <- wday_activity %>%
        select(interval, steps)

agg_wday <- selected_wday %>%
        group_by(interval) %>%
                summarise(average_steps_wday = mean(steps, na.rm = TRUE))

selected_wend <- wend_activity %>%
        select(interval, steps)

agg_wend <- selected_wend %>%
        group_by(interval) %>%
        summarise(average_steps_wend = mean(steps, na.rm = TRUE))

agg_data <- merge(agg_wday, agg_wend, by.x = "interval", by.y = "interval")

plot_wday <- ggplot(data = agg_data, aes(x = interval, y = average_steps_wday)) +
        geom_line() +
                labs(x = "Interval", y = " Average Steps", title = "Weekdays") +
                        theme(plot.title = element_text(hjust = 0.5))

plot_wend <- ggplot(data = agg_data, aes(x = interval, y = average_steps_wend)) +
        geom_line() +
                labs(x = "Interval", y = " Average Steps", title = "Weekends") +
                        theme(plot.title = element_text(hjust = 0.5))

library(gridExtra)
grid.arrange(plot_wday, plot_wend, ncol = 2)
```